---
title: "Fitting a proliferation model with complex mixture"
output:
  html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{fitting_proliferation_model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
#tools::buildVignettes(dir = ".", tangle=TRUE)
#dir.create("inst/doc")
#file.copy(dir("vignettes", full.names=TRUE), "inst/doc", overwrite=TRUE)
```

# Introduction

This vignette accompanies the ProliferationAnalysis package, to run semi
automated proliferation models in a high throughput manner. It has options to
deconvolute the effect of complex mixtures of cells where a stain negative
population is present in the sample. These stain negative cells can unduly
influence proliferation statistics.

IMPORTANT NOTE: 
This is NOT a guide to best practices to performing a proliferation experiment
and assumes some basic knowledge on getting a FACS proliferation trace ready
and evaluating its quality. This does NOT cover how to setup gates etc.

```{r setup}
library(ProliferationAnalysis)
data("ctv_example")
```


### Example data

The package comes bundled with some example traces.

#### Sample with no mixture (sample.a)
```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
sample.a <- ctv[[10]]

plot(density(log10(sample.a)),
     type="l")
```

#### Sample with high mixture (sample.b)
```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
sample.b <- ctv[[28]]

plot(density(log10(sample.b)),
     type="l")
```


# Fitting a proliferation model to CTV traces

The first step before we run proliferation modeling is to extract the raw per-cell
CTV (other dyes are available) intensities from the .fcs files and gate on single
cells \*. Depending on your goals, you can then proceed in two ways.

1. If you have no mixture of cells I would recommend gating on the CTV+ population
to avoid debris and CTV- cells affecting the fit. Then proceed with example 1.

2. If you have a complex cell sample where only a portion of cells have been
stained, proceed with example 2.

\* This package does NOT handle this part, and would refer you to the plethora of
FACS analysis options available in R. I quite like CytoExploreR for gating.

## Example one

### Step 0: Choose your optimizer

The package offers two optimization schemes.

1. [recommended] Non linear least squares (LS) with mode="LS". Here the data is first binned
and smoothed, then parameters are optimized over the smoothed trace using
non linear least squares. This method is quick and robust, but relies on binning
and some other data processing so is technically less accurate. 

2. Maximum likelihood estimation (MLE) based with mode="MLE". This mode is a
little more proper in that it does not rely on binning or other data tricks to
optimize, and takes the full data set to optimize on directly. Initial parameter
estimation is still done on a binned smoothed version of the data, but 
optimization is not. The disadvantage is it a lot slower to run, and hence I would
not recommend it. On the todo list is to speed this up, but given the least squares
works very well, this is not a high priority.

In practice I have found very little difference between these two when the
setup is performed correctly. Some examples below of issues you may encounter.

### Step 1: Identify gen0
First, Identify the rough cutoff where the undivided generation is. This cutoff needs
to be left of the mode of gen 0. I recommend putting it into the valley after
the first mode. This can be set in a consistent way for your samples using a 
CytoExploreR gate by your negative control sample. In this case we will be
setting it manually.

Note this does not directly affect the fit, and is just used to find the mode of 
gen0 in each sample.
```{r, echo=T, out.width="100%", out.width="100%", fig.dim=c(10, 6)}
gen0.cutoff <- 5.3

plot(density(log10(sample.a)), type="l")
abline(v=gen0.cutoff, lty=2)
```


### Step 2: fit peaks
In the simplest case we can run fit.peaks, which estimates peak numbers 
automatically. We will look later on at performing and tweaking a fit step by
step.
```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
# Basic fit
fit <- fit.peaks(trace=sample.a,
                 peak.0.lower.bound=gen0.cutoff,
                 mode="LS")
```

The first plot shows where the initial peak estimates are located, and how many
peaks will be optimized over. The red line indicates the mode of peak 0
the grey dotted line the lower cuttoff we set in the gen0.cutoff variable
and the grey lines the modes of subsequent peaks. The next plots show the output
of the optimization and how the initial parameters have changed during fitting.

This can be useful to diagnose issues if you get weird looking fits.

The final single plot shows the original trace vs the optimized model.


### Step 3: Evaluate and diagnose issues
If we now run a different sample, we can see there is an issue

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
# Basic fit with default parameters goes wrong
fit <- fit.peaks(trace= ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 mode="LS")
```

In this case the left tail in the data is causing the optimizer to spread out the distribution within its constraints as much as possible.

This is clearly wrong and has as a cause that the automated peak estimates don't cover this tail. In this case, there is also an extra little peak we are missing that does not pass the default peak thresholds, which is making the issue worse.

However, we can easily solve this in a couple of ways

#### Option 1: Trimming the optizier region
Enable the "auto gating" by trimming left and right tails that fall outside of
the initial model range. The initial peak estimates are first made on the full
trace that is supplied. Then the values within one peak sd outside are removed
in mode="MLE" or bins set to zero count in mode="LS". This prevents the
optimizer from fitting to the tails. The only sacrifice here is a small loss
in accuracy and potentially an issue if the initial parameter estimates are way
off, so carefully check its working as intended.

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 opt.trim.left.tail = T,
                 opt.trim.right.tail = T)
```

We can control the trimming of the tail using the options
opt.trim.right.tail and opt.trim.left.tail

In this case it makes little difference due to the way parameter constraints
are setup. But this can be sample specific!

In the following example, we first forcibly misspecify the model to show how trimming the left tail outside of the initial estimates influences the optimizer.

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
par(mfrow=c(2,1))
fit <- fit.peaks(trace=sample.a,
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0,
                 peak.max=2,
                 opt.trim.left.tail=F,
                 plot.optim=F)

fit <- fit.peaks(trace=sample.a,
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0,
                 peak.max=2,
                 opt.trim.left.tail=T,
                 plot.optim=F)
par(mfrow=c(1,1))
```

#### Option 2: Strictly gate the CTV+ region
Fix the fit by strictly gating out the on the CTV- and CTV ++ tails prior
to running.

This is a simple solution but keep in mind that it might not be the most 
accurate as with the "auto gating". The advantage over option 1 is that
here the initial peak estimates are also limited to this range.

In practice this is well within reasonable limits to do, as you would 
also do this in common software such as FlowJo or FCS express. This gating can 
easily be performed on all your samples in CytoExploreR or other FCS analysis 
packages.

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
fit <- fit.peaks(trace=ctv[[14]][log10(ctv[[14]]) > 4 & log10(ctv[[14]]) < 5.6],
                 peak.0.lower.bound=gen0.cutoff)
```

#### Option 3: Tweak the initial peak calling
You can allow more peaks to be fit in the model by relaxing initial estimation
thresholds. This can be achieved by tweaking peak.thresh.enrich, peak.thresh.summit and peak.max. To not filter any possible peaks, set these to 0.

The disadvantage here is that you might include many more peaks, increasing
model complexity, and which are not likely to be very accurate as in lower 
ranges the proliferation traces are less reliable.

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0,
                 opt.trim.right.tail=T)
```

As you can see, this produces a non-nonsensical model as well, as we have a very clear and distinct CTV- sample in this data. If we set peak.max and the the thresholds to 0 we can force only 3 peaks, or we can relax the thresholds so more peaks are considered.

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
# Force a model with 3 peaks
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0,
                 peak.max=3,
                 plot.optim = F)

# Relax the auto detect peak thresholds
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0.01,
                 opt.trim.right.tail=T,
                 plot.optim = F)
```

But the result is still not quite right! We note that this is generally a rare case for modeling peaks and should only occur if the sample was taken from a mixed cell culture. To fully fix this more complex sample, you can then either, supply opt.trim.left.tail=T with the fixed peak number or model the CTV- peak directly by supplying peak.x.model = T

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}

# Use a fixed amount of peaks
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0.01,
                 opt.trim.left.tail=T,
                 opt.trim.right.tail=T,
                 plot.optim = F)

# Explictly model a ctv- generation
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 peak.thresh.summit=0.01,
                 peak.x.model=T,
                 plot.optim = F)
```

If we now run the same parameters using MLE we will see we will find a similar
model. It just might take a little while to run
```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
fit <- fit.peaks(trace=ctv[[14]],
                 peak.0.lower.bound=gen0.cutoff,
                 mode="MLE",
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0.01,
                 opt.trim.left.tail=T,
                 opt.trim.right.tail=T,
                 plot.optim=T,
)
```

## Example Two: Modelling an extra generation

Please read example one first before moving on to this as many of the principles
are shared.

We can also choose to model an additional parameter, capturing a CTV- generation.
This additional peak is unconstrained on SD and position, which can give issues.
I'd recommend constraining the peak to be outside the CTV- gate set manually.
This only really is an issue when the CTV- population is very small,
if there is a clear peak, the autodection works well.

By specifying peak.x.model a special generation is added to the model which has
its own standard deviation and position. This is free from the constraints set
by the proliferation model. This is usefull if the trace has been done on a 
mixture of stained and unstained cells, which might bleed over into each other.

This will yield proper estimates of cell counts in later peaks where there is
bleedover, wheras gating will over-estimate cell counts.

```{r, echo=T, out.width="100%", fig.dim=c(10, 6)}
# In this case autodetection fails, as there is no clear CTV- peak, and we have not setup a constraint on where it can be
fit <- fit.peaks(trace=sample.a,
                 peak.0.lower.bound=gen0.cutoff,
                 mode="LS",
                 opt.trim.right.tail=F,
                 peak.x.model = T,
                 plot.optim=T
)

# With a bound on where the peak mean can be, in this case, there is just a spread, and no peak.x is explicitly modelled
fit <- fit.peaks(trace=sample.a,
                 peak.0.lower.bound=gen0.cutoff,
                 mode="LS",
                 opt.trim.right.tail=F,
                 peak.x.model = T,
                 peak.x.thresh.summit = 0.01,
                 peak.x.upper.bound = 3.5,
                 plot.optim=T,
)
```


In some samples the cells may have proliferated so much that there is a bleedover between divided cells and CTV- cells, or the CTV- cells have bleedover because of autofluorecence. The package can pull this apart by assuming a gausian density of the CTV- popilation.
```{r}
# Fit sample accounting for bleedover 
fit.a <- fit.peaks(trace=sample.b,
                   peak.0.lower.bound=gen0.cutoff,
                   mode="LS",
                   opt.trim.left.tail = F,
                   opt.trim.right.tail = T,
                   peak.x.model = T,
                   plot.optim=F)


get.prolif.stats(fit.a[1:7,])

# Fit sample by gating and ignoring bleedover
fit.b <- fit.peaks(trace=sample.b,
                   peak.0.lower.bound=gen0.cutoff,
                   mode="LS",
                   opt.trim.left.tail = T,
                   opt.trim.right.tail = T,
                   peak.max=6,
                   peak.x.model = F,
                   plot.optim=F)

get.prolif.stats(fit.b)

cols  <- palette.colors(n = nrow(fit.b), recycle = T)

plot(fit.a[1:7,]$peak_events,
     fit.b$peak_events,
     xlim=c(0,5000),
     ylim=c(0,5000),
     col=cols,
     pch=20,
     cex=2,
     xlab="Number of events in peak - with peak x",
     ylab="Number of events in peak - no peak x")
abline(a=0, b=1, lty=2)
legend("topright",
       legend=paste0("Gen", fit.a[1:7,]$generation),
       fill=cols)
```

# Evaluate FCS express vs this data
```{r, eval=F, echo=F}
library(minpack.lm)
library(flowCore)
#source("R/proliferation_modeling_functions.r")
#source("R/proliferation_plotting_functions.r")
devtools::install_github("https://github.com/OlivierBakker/ProliferationAnalysis/tree/main")
library(ProliferationAnalysis)
library(readxl)
source("~/Documents/tools/rscripts/plotting_functions.r")

fcs.express <- as.data.frame(read_excel("~/Documents/projects/r_packages/ProliferationAnalysisTestData/p24-27_testset/fcs_express_prolif_stats.xlsx"))
rownames(fcs.express) <- fcs.express$`Keyword: $FIL for plot #3, overlay #1`

tmp <- gsub("\\.+\\d+", "", colnames(fcs.express))
tmp <- tolower(tmp)
tmp <- gsub(" ", "_", tmp)
tmp[tmp=="keyword:_$fil_for_plot_#3,_overlay_#1"] <- "filename"
tmp <- gsub("#", "num", tmp)
tmp[12:length(tmp)] <- make.names(tmp[12:length(tmp)], unique=T)
colnames(fcs.express) <- tmp

fcs.files <- list.files("~/Documents/projects/r_packages/ProliferationAnalysisTestData/p24-27_testset/gated/", full.names = T)


peak0.valley <- 5.23 
out.stats    <- data.frame()
out.names    <- c()

i <- 1
for (file in fcs.files) {
  
  cat("[INFO] ",round((i/length(fcs.files))*100, digits=2), "%","\r")
  cur.fcs <- read.FCS(file)
  x       <- cur.fcs@exprs[,"FL1-A"]
  #plot(density(log10(x)), type="l")
  #abline(v=5.23, lty=2)
  
  if (length(x) > 500) {
    fit <- fit.peaks(x,
                   peak0.valley,
                   mode="LS",
                   plot=F)
    stats <- get.prolif.stats(fit)
    
    out.stats <- rbind(out.stats, stats)
    out.names <- c(out.names, basename(file))
  }

  i <- i+1
}

rownames(out.stats) <- out.names
colnames(out.stats) <- names(stats)

# Plot of counts
tx <- as.numeric(fcs.express[basename(file), grep("num_of_cells", colnames(fcs.express))])
ty <- fit$peak_events
p1 <- xy.plot(tx[1:length(ty)], 
        ty,
        fixed=T,
        xlab="Number of cells per peak FCS express",
        ylab="Number of cells per peak R",
        size=4)
theme.plain(p1)


ol <- intersect(rownames(fcs.express), rownames(out.stats))

p1 <- xy.plot(fcs.express[ol,]$division_index,
        out.stats[ol,]$division_index,
        xlab="FCS express division index",
        ylab="R division index",
        fixed=T)


theme.plain(p1)

p1 <- xy.plot(fcs.express[ol,]$proliferation_index,
        out.stats[ol,]$proliferation_index,
        xlab="FCS express proliferation index",
        ylab="R proliferation index",
        fixed=T)


theme.plain(p1)

```



```{r, eval=F, echo=F}

# Debug

library(ProliferationAnalysis)
source("R/proliferation_modeling_functions.r")
source("R/proliferation_plotting_functions.r")

# Simulate proliferation data #2
y <- 10 ^ rnorm(1000, mean=10, sd=0.05)
y <- c((y/8)[1:500], y/4, y/2, y[1:500])

peaks <- fit.peaks(y, peak.0.lower.bound=9.9, mode="LS", full.out=T)


dens <- prolif.model.density(peaks$peak.stats$opt_mean,
                     peaks$peak.stats$opt_peak_sd,
                     peaks$peak.stats$opt_summit,
                     peaks$x,
                     log=T)

nll <- prolif.model.nll(peaks$peak.stats$opt_mean,
                     peaks$peak.stats$opt_peak_sd,
                     peaks$peak.stats$opt_summit,
                     peaks$x,
                     log=T)


plot(peaks$x, exp(dens))

peaks <- fit.peaks(y, peak.0.lower.bound=9.9, mode="MLE", full.out=T, verbose=T)




# Development area
library(ProliferationAnalysis)
data("ctv_example")
source("R/proliferation_modeling_functions.r")
source("R/proliferation_plotting_functions.r")
library(minpack.lm)
library(LaplacesDemon)
sample.b <- ctv[[28]]
sample.b <- ctv[[36]]


sample.b <- ctv[[10]]
sample.b <- sample.b[log10(sample.b)>2]
gen0.cutoff <- 5.3


sample <- sample.int(length(sample.b), 2000)
system.time({
fit2 <- fit.peaks(trace=sample.b[sample],
                 peak.0.lower.bound=gen0.cutoff,
                 mode="MLE",
                 opt.trim.left.tail = F,
                 opt.trim.right.tail = T,
                 peak.x.model = T,
                 peak.x.upper.bound = 3.8,
                 peak.thresh.enrich=0,
                 peak.thresh.summit=0,
                 peak.max=10,
                 full.out=T,
                 verbose=F,
                 log=F,
                 bins=1000)
})


yy <- prolif.model.density(fit$peak.stats$est_mean,
                 fit$peak.stats$est_peak_sd,
                 fit$peak.stats$est_summit_percentage,
                 fit$x,
                 log=T,
                 verbose=T)




binwidth      <- fit$x[2] -fit$x[1]
total.events  <- sum(fit$y)

library(mixtools)

tmp <- normalmixEM(log10(sample.b),
                   lambda=fit$peak.stats$est_summit_percentage,
                   mu=fit$peak.stats$est_mean,
                   sigma=fit$peak.stats$est_peak_sd,
                   mean.constr=fit$peak.stats$est_mean,
                   sd.constr=c(rep("a", nrow(fit$peak.stats)-1), NA))

yy <- prolif.model.density(tmp$mu,
                 tmp$sigma,
                 tmp$lambda,
                 fit$x,
                 log=F,
                 verbose=T)

yy.plot       <- (yy*binwidth) * total.events

plot(fit$x, fit$hist$counts, pch=20, col="grey")
lines(fit$x, yy.plot, col="blue")




yy <- prolif.model.density(fit.mle$peak.stats$opt_mean,
                 fit.mle$peak.stats$opt_peak_sd,
                 fit.mle$peak.stats$peak_area,
                 fit.mle$x,
                 log=F,
                 verbose=T)

yy.plot       <- (yy*binwidth) * total.events

plot(fit$x, fit$hist$counts, pch=20, col="grey")
lines(fit$x, yy.plot, col="blue")

yy <- prolif.model(fit.mle$peak.stats$opt_mean,
                 fit.mle$peak.stats$opt_peak_sd,
                 fit.mle$peak.stats$opt_summit / fit.mle$peak.stats$opt_peak_sd,
                 fit.mle$x)
yy.plot       <- (yy*binwidth/2) * total.events


plot(fit$x, fit$hist$counts, pch=20, col="grey")
lines(x, yy.plot, col="blue")
#-----------------------------------------------------

plot(fit$peak.stats$peak_area_prop, fit.mle$peak.stats$peak_area_prop)
abline(a=0, b=1)

plot(fit$peak.stats$peak_events, fit.mle$peak.stats$peak_events)
abline(a=0, b=1)


get.prolif.stats(fit$peak.stats[1:(nrow(fit$peak.stats)-1),])
get.prolif.stats(fit.mle$peak.stats[1:(nrow(fit$peak.stats)-1),])

#-----------------------------------------------------
# Development area
library(ProliferationAnalysis)
data("ctv_example")
source("R/proliferation_modeling_functions.r")
source("R/proliferation_plotting_functions.r")

sample.b <- ctv[[28]]
sample.b <- ctv[[10]]
gen0.cutoff <- 5.3
trace               <- sample.b
peak.0.lower.bound  <- 5.3
bins                <- 250
smoothing.window    <- 2
plot.optim          <- T
plot.final          <- T
plot                <- T
plot.main           <- "Proliferation model"
opt.peak.pos.dev    <- NULL
opt.trim.left.tail  <- F
opt.trim.right.tail <- T
mode                <- "LS"
full.out            <- F
verbose             <- F


peak.x.model             <- T
peak.x.upper.bound       <- 4  # Optional hard limit for peak x
peak.x.thresh.summit     <- 0.05 # Peak summit must be at least 5% of the data mode
peak.x.thresh.enrich     <- 1.2  # threshold for stopping


peak.x.mode <- find.peak.x.approx.mode(log10(sample.b),
                                       peak.x.thresh.summit=peak.x.thresh.summit,
                                       peak.x.thresh.enrich=peak.x.thresh.enrich,
                                       peak.x.upper.bound=peak.x.upper.bound)

plot(density(log10(sample.b)), type="l")
abline(v=peak.0.lower.bound, lty=2)
abline(v=peak.x.upper.bound, lty=2)
abline(v=peak.x.mode, lty=2, col="blue")

```



